Visual narratives of the Covid-19 pandemic

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
update on paper and references

%% - \Abstract{} almost as usual
\Abstract{
  Covid-19 has created a world-wide interest in understanding the dynamics of a pandemic. Hence, numerous media outlets as well as researchers have produced comprehensive data visualizations to illustrate the relevant trends and figures. In this paper, we will look at an elective choice of Covid-19 data visualizations to evaluate and discuss currently established visualization tools in their capacity to provide a communication channel both within the data science team and also between data analyst, domain experts and a more general interested audience. While there is no fixed catalogue of evaluation criteria for data visualizations we will try to provide an overview on the different core aspects of visualization evaluation and their competing principles.
  }

@@

Here some general introduction

Effective participation in the social dynamics of a democracy requires clear understanding of quantitative evidence and rules by the people. As the medical community, governments, business, and the public come together to fight the COVID-19 pandemic, scientists have a special role in helping to communicate what the data actually mean.
For data analysis and modeling to have an impact as a component of decision making, appropriate reporting and communication is key. There are numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). These standards are based on commonly accepted core quality principles and values such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. For measures to restrain and overcome an epidemic effectively, communication among experts that follows highest professional and ethical standards is not sufficient. In a democratic society, policy measures can only be implemented if they are based on the acceptance of the wider population. This puts high demands on skills associated with communicating statistical evidence on the side of scientists, governments and media, and a citizenry able to understand statistical messages. %

In recent decades, there have been numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information, see \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}, to name only a few. Data journalism has recently taken off as an innovative component of news publishing, and COVID-19 provides numerous excellent examples, often using an interactive visual format on the Internet, such as dashboards. %Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges. 
A fundamental problem in assessing probabilities lies, for example, in the intuitive conflation of subjective risks ("how likely am I to become infected") and general risks ("how likely is it that some person will become infected"). Another issue is that of equating sensitivity of a diagnostic test and the positive predictive value \citep{Eddy1982,Gigerenzer2007,McDowell2019,Binder2020}. In particular, the prevalence (or base rate) is often neglected leading to this confusion. As \cite{Hoffrage} {\color{blue}Hoffrage et al. (2000)} point out “… statistics expressed as natural frequencies improve the statistical thinking of experts and nonexperts alike”. Smooth sliding back and forth between probabilities and natural frequencies is easy once figures are related to a unifying reference value. A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%. Using absolute in place of relative numbers, representing information in appealing graphic forms, and summarizing most relevant facts in “fact boxes” are some of the methods developed and advocated for increasing transparency in communication between stakeholders such as health providers and patients (See, for example: https:// www. hardingcenter.de)
Fact boxes combined with icon arrays are recommended for the presentation of test results. Both representations are based on natural frequencies \citep{[62],Krauss2020} and present case numbers as simply and concretely as possible. Many scientific studies show that icon arrays help people understand numbers and risks more easily (e.g. \cite{McDowell2019}). The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.

@@

% May also want to mention at the outset the limitations in the data due to low testing capacity, weekly effects due to reporting, and so on. We must be able to represent the data that exist visually while acknowledging the limitations associated with that data.

@@

Visual representations take a central position in public communication and aim to represent the corresponding dynamics and contents in a quickly understandable way. Usually either time-dependent parameters or data with a spatial reference are visualised. For spatially distributed data, choropleth maps are predominantly used, in which administrative regions defined by the responsible health authorities are coloured according to the distribution density of the infection figures or variables derived from them (see Figure~\ref{fig:choro}). Their visual perception problems - such as the visual dominance of the area of administrative regulatory frameworks that have no direct relation to infection events - are well known but still widespread.  In addition, the use of ordinance thresholds as the basis for color scaling is often at odds with color schemes that emphasize real spatial distributional differences.

\begin{figure*}
	\includegraphics[width = 0.9\textwidth]{choroplethen}
	\caption{Choropleth map of the incidence figures for Germany by district. Source: Robert-Koch-Institute https://app.23degrees.io/export/oCRP768wQ3mCswE7-choro-corona-faelle-pro-100-000/image.}
	\label{fig:choro}
\end{figure*}

For time-dependent parameters, different variants of time series diagrams are used, predominantly line and column diagrams. Classic errors of graphical representation, such as overemphasising temporal variability by reducing the value axis to a small section, have now largely disappeared from the media. Switching between line and bar charts for purely design-related reasons in order to produce corresponding graphical diversity nevertheless seems questionable. 
The use of logarithmic scales in time series diagrams should be evaluated with caution. On the one hand, they tempt superficial readers to underestimate dynamic growth processes; on the other hand, they increase the demands on the mathematical and statistical literacy of the readership without corresponding advantages of visual representation. Figure~\ref{fig:incidence} shows the time course of the 7-day incidence per 100,000 people between 24 January and 4 February 2021 for some selected countries. While the differences appear relatively small on the logarithmic scale, the linear scale shows considerable differences.

\begin{figure}
		\includegraphics[width = 0.49 \textwidth]{incidence1.png}
			\includegraphics[width = 0.49 \textwidth]{incidence2.png}
	\caption{The 7-day incidence for different countries over the course of time. On the logarithmic scale (left graphic), differences seem small. The linear scale (right graphic), however, shows considerable differences. Source: Our World in Data, https://ourworldindata.org/covid-cases?country=IND~USA~GBR~CAN~DEU~FRA.}
	\label{fig:incidence}
\end{figure}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
  Covid-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules. For this reason, numerous media and research groups have produced comprehensive data visualisations to illustrate the relevant trends and figures. In this paper, we will look at a selection of Covid 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience. Although there is no set catalogue of evaluation criteria for data visualisations, we will try to give an overview of the different core aspects of visualisation evaluation and their competing principles.
  }

@@

Over the past two years, several waves of Covid-19 infections with different mutants of the SARS-CoV-2 virus have swept across the globe, claiming many lives, causing numerous health damages and affecting our personal lives in many ways. According to the WHO, over 304 million confirmed cases and over 5.4 million deaths have been reported as of early January 2022 \footnote{\url{https://www.who.int/publications/m/item/weekly-epidemiological-update-on-covid-19---11-january-2022}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. So there are several goals to be achieved by visualising Covid-19 data. These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace. But as with any other data visualisation two general principles remain the same: ensuring clear vision by optimising the data-to-ink ratio \citep{Tufte2001} (or signal-to noise ratio) and ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic as they have been showcasing their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an innovative part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards.

At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

The Covid-19 pandemic has provided ample evidence that policy measures are only effectively implemented and adhered to by the people if they are accepted by a vast majority of the population. To achieve this goal, effective communication of quantitative evidence and statistical results among scientists, governments, media, and the citizens is essential to justify the adequacy, usefulness, and relevance of the measures.

A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%. Using absolute in place of relative numbers, representing information in appealing graphic forms, and summarizing most relevant facts in “fact boxes” are some of the methods developed and advocated for increasing transparency in communication between stakeholders such as health providers and patients (See, for example: https:// www. hardingcenter.de)
Fact boxes combined with icon arrays are recommended for the presentation of test results. Both representations are based on natural frequencies \citep{[62],Krauss2020} and present case numbers as simply and concretely as possible. Many scientific studies show that icon arrays help people understand numbers and risks more easily (e.g. \cite{McDowell2019}). The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.

Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges.

@@

@@

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global perspective of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example Figure~\ref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{who_totalcases_choro.png}
	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
	\label{fig:choro1}
\end{figure*}
